Bayesian Detection of Bias in Peremptory Challenges Using Historical Strike Data

### 2022-09-30 analysis/paper_draft3.Rmd
added text on Batson procedure

Following @Kadane2018a and @Barrett2007 [see also @Gastwirth2005, 183], we model a peremptory-challenge process in which each party strikes prospective jurors in an alternating sequence. Under such a procedure, the trial judge rules on all challenges for cause before the parties exercise any peremptories. Then, of the potential jurors who remain, a subset of them are subject to peremptory strike, usually a number that corresponds to the number seats on the jury (plus alternates, if any). The parties exercise their strikes on anyone among this subset of potential jurors in an alternating sequence. Once all strikes are used or waived, the remaining prospective jurors are each assigned to seats on the jury plus a few alternate jurors.

@@

Moreover, with a tool like our prototype app, attorneys can use the approach here in to detect strike bias for _Batson_ and related challenges, in three ways. First, the posterior distribution of the bias parameter can help the lawyer decide in real time whether it is worth raising a  _Batson_ challenge at all. Second, it can help the _Batson_ challenger meet the initial burden of producing just enough evidence of strike bias to persuade the trial judge to require the striking attorney to provide (permissible) reasons for their challenged strikes. Third, once the striking attorney proffers those reasons, the trial judge can consider the attorney's estimated bias parameter as relevant to how much to take those proffered reasons as what the attorney actually believed as well as whether the alleged illegal strike bias is ultimately more likely than not to be true.

>>>>

Following others (e.g., @Kadane2018a; @Barrett2007; @Gastwirth2005, 183], we model a peremptory-challenge process in which each party strikes prospective jurors in an alternating sequence. Under such a procedure, the trial judge rules on all challenges for cause before the parties exercise any peremptories. Then, of the potential jurors who remain, a subset of them are subject to peremptory strike, usually a number that corresponds to the number seats on the jury (plus alternates, if any). The parties exercise their strikes on anyone among this subset of potential jurors in an alternating sequence. Once all strikes are used or waived, the remaining prospective jurors are each assigned to seats on the jury plus a few alternate jurors.

@@

Moreover, with a tool like our prototype app, attorneys can use the approach here in to detect strike bias for _Batson_ and related challenges, in three ways. First, the posterior distribution of the bias parameter can help the lawyer decide in real time whether it is worth raising a  _Batson_ challenge at all.

```{r batsonworkflowfig}

library(DiagrammeR)
grViz(diagram = "digraph flowchart {
  rankdir = LR
  node [fontname = arial, shape = box]
  tab1 [label = '@@1']
  tab2 [label = '@@2']
  tab3 [label = '@@3']

  tab1 -> tab2 -> tab3;
}

  [1]: 'prima facie case'
  [2]: 'proffered reasons'    
  [3]: 'prove bias'    
  ", height = '100%', width = '100%')

Second, if raised, the tool can help the _Batson_ challenger meet its evidentiary burdens at key steps of a _Batson_ challenge. A _Batson_ challenge typically proceeds in a three-step sequence (Figure \@ref(fig:batsonworkflowfig)). First, the challenger must meet an initial burden (the _prima facie_ case) of producing just enough evidence of strike bias to raise a sufficient inference of illegal bias. If so, then at step two, the striking attorney must produce (permissible) reasons for their challenged strikes. Third, the challenger must discredit those proffered reasons and ultimately persuade the trial judge to find that it is more likely than not that the striking attorney acted with illegal bias.

To meets its evidentiary burdens at the first and third steps, the attorney can introduce the striking attorney's estimated bias parameter as relevant statistical evidence. In this way, the tool helps detect bias. But like any statistical evidence submitted for a _Batson_ challenge, it cannot alone be the basis of deciding a _Batson_ challenge. That is because it alone cannot account for the veracity of the striking attorney's reasons offered at step two and because the _Batson_ procedure allows for any type of circumstantial evidence of bias at steps one and three.
